The Hot Hand Fallacy in Call of Duty Search and Destroy

### 2022-08-25 paper/paper.tex
remove gt code, rename constant vars, proof-read up until result section, rate -> proportion in prose

\begin{abstract}
Our research investigates patterns in round win percentages in
professional Search and Destroy (SnD) matches of the popular
first-person shooter game Call of Duty (CoD). First, we find evidence
suppOrting the hypothesis that round win probability can be modeled as a
constant across rounds in the series, although not at the naive 50\%.
Second, we examine post-streak round win probability, given the series
outcome. Given streak length, series length, and series winner, we find
no evidence that post-streak win rate is significantly different than
expected, suggesting that the perception of momentum at the team level
is deceiving. Wald-Wolfowitz run tests also fail to provide evidence for
the hot hand phenomenon.
\end{abstract}

@@

Call of Duty (CoD), first released in 2003, is one of the most popular
first-person shooter (FPS) video game franchises of all-time. The most
popular mode in the competitive scene is ``Search and Destroy'' (SnD),
in which one team tries to destroy one of two designated bomb sites on
the map, defended by the opposing team.\footnote{SnD bears resemblance
  to ``Bomb Defusal'' in Counter-Strike and ``Plant/Defuse'' in
  Valorant, two other FPS games played in more popular professional
  leagues.}

In professional CoD SnD, the two teams\footnote{In the 2020 season,
  teams played with five players each; in the 2021 and 2022 seasons,
  teams played with four players each.} take turns playing offense and
defense every round. The first to six round wins (best-of-11 format) is
declared the series winner.\footnote{Each round has a two-minute time
  limit, not counting the potential extension granted if a bomb is
  planted.} A round can end in one of four ways:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  One team eliminates all members of the other team prior to a bomb
  plant. (Eliminating team wins.)
\item
  The offensive team eliminates all members of the defensive team after
  a bomb plant.\footnote{\begin{itemize}
    \tightlist
    \item
      The bomb can be picked up by any member of the offensive team.
    \item
      The bomb carrier is not obstructed by carrying the bomb
      (i.e.~movement is the same, weapon usage is the same, etc.).
    \item
      The defense does not get any visual indication for who is carrying
      the bomb.
    \item
      A bomb plant takes five seconds. The timer resets if the player
      stops planting site prior to completing it.
    \item
      A bomb defuse takes seven seconds. The timer resets if the player
      ``drops'' the bomb.
    \item
      Once the bomb is planted, the round timer is reset to 45 seconds.
    \end{itemize}} (Offense wins.)
\item
  The defense defuses the bomb after a bomb plant.\footnote{Often the
    defensive team will try to eliminate all team members prior to
    making the defuse, but in some cases, they may try to ``ninja''
    defuse.} (Defense wins.)
\item
  The offense does not make a plant by the time the round timer ends.
  (Defense wins.)
\end{enumerate}

We adopt the terminology ``series'' to refer to a single best-of-11
matchup, so as to mirror the terminology of playoff series in
professional leagues like the National Basketball Association (NBA),
National Hockey League (NHL), and Major League Baseball (MLB). A
``game'' or a ``match'' in such leagues is analogous to a ``round'' of
CoD SnD for the purposes of this paper.\footnote{Further, these are
  analogous to a trial in statistical analysis of Bernoulli events.}

SnD is not the only game type played in competitive CoD. Teams play in a
head-to-head, best-of-five format, where SnD is always played as the
second and fifth game ``maps'', as most players refer to them (if a
fifth map is needed). The best-of-five matchup could also be called a
``series'', but since we analyze only the SnD game types, we refer to
SnD games as series.

@@

The data set consists of all SnD matches played in major tournaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852
series.\footnote{288 of the series occur in major tournaments, which are
  considered to be more ``competitive'' than the qualifiers since they
  are played in person (COVID-permitting), whereas the qualifiers are
  played online.} Data has been collected in spreadsheets by community
member ``IOUTurtle''.\footnote{Raw data:
  \url{https://linktr.ee/CDLArchive}. Author:
  \url{https://twitter.com/IOUTurtle}. Processed data can be found at
  \url{https://github.com/tonyelhabr/fps_round_games/blob/master/data/cod_rounds.csv}.}

6-0 sweeps make up 4.7\% of all series, while 6-5 series make up 23.7\%
of series. The observed offensive round win percentage across all rounds
is \(\tau_O\) = 47.8\%.\footnote{Offensive round win percentage has been
  nearly constant across the three games during the CDL era: 1. 47.2\%
  in MW (2020) 2. 47.9\% in Cold War (2021) 3. 48.1\% in Vanguard (2022)}
Table \ref{tbl:o-win-prop-by-series-state} shows round win percentages
by series ``state'' (i.e.~the number of round wins by each team prior to
an upcoming round). Offensive round win rate is not quite constant,
although never veers more than 10\% from \(\tau_O\).

@@

At tournaments, higher-seeded (``better'') teams choose whether they
want to start on offense in the SnD series played as the second or fifth
map, and their opponent is assigned to start on offense in the SnD map
not chosen.\footnote{Hardpoint has been played as the first and fourth
  maps in a matchup. The variant for the third map was Domination in the
  2020 season, and Control in the 2021 and 2022 seasons.}\footnote{Anecdotally,
  the higher-seeded teams tend to choose to play defense in the first
  round of the SnD series played as the second game, although this
  choice is not consistent across teams, or with the same team over
  time.} Outside of tournaments, teams all play each other twice
throughout the season, constituting qualifiers. Each team gets to play
the higher-seeded role once in their head-to-head matchups with a given
team, despite their win-loss records. Overall, we can argue that
potential bias due to stronger teams playing defense---or, if offense,
if offense happens to be generally advantageous for a given map---is
very minimal.

@@

There have been a handful of studies of the distribution of games played
in a series of a professional sport. Most assume a constant probability
\(\phi\) of a given team winning a game in the series, regardless of the
series state. Mosteller (1952) observed that the American League had
dominated the National League in MLB World Series matchups through 1952,
implying that games should not be modeled with a constant (i.e.~the
null) \(\phi_0 = 0.5\). Mosteller proposed three approaches for
identifying the optimal constant probability value of the stronger team
in the World Series, finding \(\phi \approx 0.65\) in each case: (1)
solving for \(\phi\) from the observed average number of games won by
the loser of the series, which he called the ``method of moments''
approach; (2) maximizing the likelihood that the sample would have been
drawn from a population in which the probability of a team winning a
game is constant across the series (i.e.~maximum likelihood), and (3)
minimizing the chi-squared goodness of fit statistic \(\chi^2\) as a
function of \(\phi\).

Chance (2020) re-examines the constant probability notion in the MLB
World Series (1923--2018), the NBA Finals (1951--2018), and the NHL
Stanley Cup (1939--2018). Chance found strong evidence against the null
hypothesis of \(\phi_0 = 0.5\) in the MLB and NHL championship series
when applying Mosteller's first and second methods.

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use ``streaks'' and momentum
  interchangeably, but as Steeger et al.~((2021)) note, momentum implies
  dependence between events, whereas streaking does not.} Two opposing
fallacies are observed in the context of momentum: the ``gambler's
fallacy'' (negative recency) and the ``hot hand fallacy'' (positive
recency). Per Ayton et al. (2004), negative recency is ``the belief
that, for random events, runs of a particular outcome \ldots{} will be
balanced by a tendency for the opposite outcome'', while positive
recency is the expectation of observing future results that match recent
results.

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. Gilovich et al. (1985),
henceforth GVT, ido not find evidence for the hot hand phenomenon.
However, Miller and Sanjurjo (2018), henceforth MS, provided a framework
for quantifying streak selection bias, which effectively works in the
manner posited by the gambler's fallacy. Specifically, MS say that a
``bias exists in a common measure of the conditional dependence of
present outcomes on streaks of past outcomes in sequential data''
implying that, under i.i.d. conditions, ``the proportion of successes
among the outcomes that immediately follow a streak of consecutive
successes is expected to be strictly less than the underlying
(conditional) probability of success''. When applying streak selection
bias to GVT's data, MS came to the opposite conclusions as GVT.

Other research has borrowed a common approach from the field of quality
control, looking at unlikely sequences of events with the Wald-Wolfowitz
runs test (Peel and Clauset 2015, Steeger et al.~2021). Peel and Clauset
found no evidence for unlikely sequences of scoring events in the NHL,
College Football (CFB), and National Football League (NFL), but did in
the NBA. As a check on their entropy approach to momentum
identification, Steeger et al.~found several NHL teams with sequence of
wins in the 2018-2019 regular season violating the Wald-Wolfowitz null
hypothesis.

@@

Despite the plethora of existing research on games played in a series
and momentum in sports, these topics have yet to be investigated
heavily, if at all, in esports. Work has been done to examine
intra-round win probability in other FPS titles such as Counter-Strike
(Xenopoulos, Freeman, and Silva 2022) and Valorant (DeRover 2021), both
of which are round-based like CoD SnD. However, considering
Counter-Strike and Valorant specifically, research on round-level trends
seems non-existent, perhaps for one of several reasons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Both have economic aspects that can create clear advantages for one
  team in a round, given how prior rounds played out. CoD has no such
  equivalent, except for perhaps ``score streaks'', which infrequently
  occur.
\item
  Teams play either offense or defense for many consecutive rounds. (The
  former has a 15-15-1 format and the latter uses a 12-12-1 format.) On
  the other hand, teams in CoD SnD rotate roles every round, analogous
  to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series
  for professional sports like the MLB, NBA, and NHL.\footnote{1-1-1-1-1-1-1
    is not used today in these leagues, but it was at least once in each
    league.} While theoretically one might be able to account for any
  kind of format, such as a 5-5-1, the rotation of team sides every
  round is convenient for convincing ourselves that rounds could
  reasonably be modeled as i.i.d. Bernoulli trials.
\item
  Both Counter-Strike and Valorant have overtime rules---team must win
  by two rounds---which can make end-of-series sequences difficult to
  model. CoD SnD does not have overtime rules.
\end{enumerate}

To our knowledge, there is no existing public statistical research on
the CoD SnD format, beyond descriptive analysis on social media. While
intra-round trends may be more directly applicable to teams looking for
an advantage on their competition, broader investigation of a concept
like the hot-hand fallacy in a different sport---particularly one that
is less subjected to factors that may be difficult to control for,
e.g.~weather---should be useful as a precedent or supplement for future
researchers.

@@

To evaluate the constant round win probability null hypothesis---that
is, that the expected and observed round win rates, \(\hat{\Phi}(r)\)
and \(\Phi(r)\) respectively, are equal to one another---we can compute
the chi-square goodness of fit statistic

\begin{equation}\label{eq:chi-squ}
\chi^2 = \sum_{r \in R} \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)},
\end{equation}

in which \(R = [6, 7, 8, 9, 10, 11]\) for CoD SnD.

@@

Let us now consider round win rate immediately following a streak of
\(k\) wins, given that the series lasts \(r\) rounds. Removing our
knowledge of a streak, we might model the Bernoulli round win
probability as

@@

As shown by MS with their Theorem 1, we should expect the proportion of
round wins immediately following a streak of \(k\) rounds wins for a
series lasting \(r\) rounds, \(\hat{P}^{+|k,r}_{MS}\), to be strictly
less than \(p^{+|r}\).\footnote{See Appendix E.1 of MS for the proof.}
Unfortunately, there does not exist a closed form representation for the
expected value of \(\hat{P}^{+|k,r}_{MS}\) for \(k > 1\). Nonetheless,
one may run simulations to estimate the expected value, as we choose to
do.\footnote{We have adapted the code from Vafa (2017), which implements
  MS's framework (2018).}

Although one might be tempted to mirror the hypothesis testing for
difference of proportions performed by MS (and GVT), our context is
fundamentally different from that of MS, who focus on longitudinal data
in controlled settings.\footnote{GVT also perform statistical tests on
  shots from players in live games, i.e.~``observational'' data, but
  they note that their findings are likely affected player shot
  selection in the face of defensive strategy by the opposing team.} The
number of trials is fixed in their experimental designs, but in CoD SnD,
the number of rounds played is determined as a function of the max
number of possible rounds (\(s\)) and whether or not the team wins the
series. The Bernoulli trial success probability, i.e.~the single round
win rate in CoD SnD, is not independent of the opponent. Consequently, a
statistical test of the difference in \(\hat{P}^{+|k,r}\) and
\(\hat{P}^{-|kr}\), as performed by MS to evaluate their hypothesis
regarding post-streak success rate, is not completely appropriate,
although useful as a reference.

@@

We can perform a binomial test to evaluate the null hypothesis
\(\omega = \omega_0\) for the observed probability of success \(\omega\)
and a user-specified \(\omega_0\) (where \(0 \leq \omega_0 \leq 1\)). If
there are \(r^+\) observed successes in a sample of \(r\) trials and we
expect that there should be \(r * \omega_0\), the probability of
arriving at this expected number of successes is

@@

Treating the notional proportion \(\hat{P}^{+|k,r}_0\) as the null
\(\omega_0\) and plugging in the observed proportion \(P^{+|k,r}\) for
\(\omega\) in Equation \ref{eq:binom} (treating proportions as
probabilities), we can evaluate the null hypothesis that the observed
probability is equal to the notional probability. If we can reject this
null hypothesis, then we can consider team momentum, represented by
post-streak success, plausible.

One can perform the same binomial test for the MS's
streak-selection-adjusted proportion,
\(\hat{P}^{+|k,r}_{MS}\).\footnote{Given the caveats mentioned before,
  these results should be heeded with caution.}

@@

Again, we caapply a binomial test to evaluate the hypothesis that the
expected proportions, \(\hat{P}^{+|k,r,i}_0\) and
\(\hat{P}^{+|k,r,i}_{MS}\) separately, are equal to the observed
proportion \(P^{+|k,r}\).

@@

Stepping back from the one-sided perspective of a single team's round
win probability when streaking, one can attempt to detect the hot hand
phenomenon with a Wald-Wolfowitz runs test. Under the null hypothesis,
the number of runs in a sequence of \(r\) trials, \(\zeta(r)\), is a
random variable that can take on values \(r^+\) and \(r^- = r - r^+\)
for the number of success and failure respectively, with the following
mean \(\mu\) and variance \(\sigma^2\):

@@

One can subset the observed series sequences to those that violate the
null hypothesis for the run test and perform a test of equal
proportions, where the null is that the observed proportion of a
sequence relative to all possible sequences, \(P(\zeta) = P^\zeta\), is
equal to the expected proportion of the sequence relative to all
possible sequences, \(\hat{P}^\zeta\). The test statistic is

\begin{equation}\label{eq:prop}
Z = \frac{P^\zeta - \hat{P}^\zeta}{\sqrt{P^\zeta_{\delta} (1 - P^\zeta_{\delta}) (1 / N^\zeta + 1 / \hat{N}^\zeta)  } }
\end{equation}

@@

and where \(N^\zeta\) is the observed number of sequences,
\(\hat{N}^\zeta\) is the expected number of sequences. If we can reject
the null hypothesis for such sequences, then there is implicit evidence
in favor of the hot hand.

@@

First, we investigate the constant probability assumption and the
distribution of rounds played in a series. Chance's work is closely
related to our, and, in fact, provides a guide for this investigation.
Afterwards, We investigate post-streak win rates, referring to work from
MS. We finish with a brief look at sequences of round wins violating the
Wald-Wolfowitz runs test, and whether such observed sequences occur more
frequently than expected.

@@

Table \ref{tbl:mosteller-methods-results} shows the alternate values for
the constant round win probability that we find when applying the three
methods suggested by Mosteller (1952). Each is approximately equal to
0.575. When applying Equation \ref{eq:chi-squ}, each results in a
\(\chi^2\) value for which we cannot reject the constant probability
null hypothesis.

@@

Observing that \(\hat{\Phi}_2(r)\) reasonably matches \(\Phi(r)\)
(especially in comparison to \(\hat{\Phi}_0(r)\)), along with the null
hypothesis rejection shown in Table \ref{tbl:mosteller-methods-results},
we can say that the constant round win probability assumption is valid
in CoD SnD series with the appropriate choice of \(\phi\)
(\(\approx 0.575\)).

@@

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), we focus
on streaks of three round wins.\footnote{Three happens to also be a
  reasonable number for series that last at maximum 11 rounds.} Table
\ref{tbl:pwkr} compares the notional and MS expected proportions,
\(\hat{P}^{+|k,r}_0\) and \(\hat{P}^{+|k,r}_{MS}\) respectively, with
the observed round win rate, \(P^{+|kr}\), following streaks of \(k=3\)
round wins given the series outcome.

@@

With the exception of \(\hat{P}^{+|k,r}_{MS}\) when \(r = 10\) and the
eventual series loser is the team that wins after a streak of three
round wins, all un-adjusted binomial test \(p\)-values are greater than
the \(\alpha = 0.05\) confidence level, implying that we cannot reject
the null hypothesis that the expected notional and MS post-streak round
win rates are different than the observed proportion. However, after
adjusting the \(p\)-values with the Benjamini and Yekutieli (BY)
correction (2001), the null hypothesis cannot be rejected for any
case.\footnote{As noted before, the MS expectations may be unreliable
  for CoD SnD, so one is inclined to prefer the results of the notional
  binomial tests.}

@@

However, our results do not provide evidence for momentum on several
fronts: (1) the evidence supporting the constant round probability; (2)
the failure to reject the binomial null hypothesis for post-streak win
rates; and (3) the failure to reject the equal proportions null
hypothesis for streaky round sequences identified by the Wald-Wolfowitz
runs test.

>>>>

\begin{abstract}
Our research investigates patterns in round win percentages in
professional Search and Destroy (SnD) matches of the popular
first-person shooter game Call of Duty (CoD). First, we find evidence
supporting the hypothesis that round win probability can be modeled as a
constant across rounds in the series, although not at the naive 50\%.
Second, we examine the proportion of round wins immediately following a
win streak. Given streak length, series length, and series winner, we
find no evidence that post-streak win proportion is significantly
different than expected, suggesting that the perception of momentum at
the team level is deceiving. Wald-Wolfowitz run tests also fail to
provide evidence for the hot hand phenomenon.
\end{abstract}

@@

Call of Duty (CoD), first released in 2003, is one of the most popular
first-person shooter (FPS) video game franchises of all-time. The most
popular mode in the competitive scene is ``Search and Destroy'' (SnD),
in which one team tries to destroy one of two designated bomb sites on
the map.\footnote{SnD bears resemblance to ``Bomb Defusal'' in
  Counter-Strike and ``Plant/Defuse'' in Valorant, two other FPS games
  played in more popular professional leagues.}

In professional CoD SnD, the two teams\footnote{In the 2020 season,
  teams played with five players each; in the 2021 and 2022 seasons,
  teams played with four players each.} take turns playing offense and
defense every round. The first to six round wins (best-of-11 round
format) is declared the series winner.\footnote{A short list of rules
  that govern timing follows.

  \begin{itemize}
  \tightlist
  \item
    Each round has a 90 second time limit, not counting the potential
    extension granted if a bomb is planted
  \item
    A bomb plant takes five seconds. The timer resets if the player
    stops planting site prior to completing it.
  \item
    A bomb defuse takes seven seconds. The timer resets if the player
    ``drops'' the bomb.
  \item
    Once the bomb is planted, the round timer is set to 45 seconds.
  \end{itemize}} A round can end in one of four ways:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  One team eliminates all members of the other team prior to a bomb
  plant. (Eliminating team wins.)
\item
  The offensive team eliminates all members of the defensive team after
  a bomb plant. (Offense wins.)
\item
  The defense defuses the bomb after a bomb plant.\footnote{Often the
    defensive team will try to eliminate all team members prior to
    making the defuse, but in some cases, they may try to ``ninja''
    defuse.} (Defense wins.)
\item
  The offense does not make a plant by the time the round timer ends.
  (Defense wins.)
\end{enumerate}

We adopt the terminology ``series'' to refer to a single best-of-11
matchup, so as to mirror the terminology of playoff series in
professional leagues like the National Basketball Association (NBA),
National Hockey League (NHL), and Major League Baseball (MLB). A
``game'' or a ``match'' in such leagues is analogous to a ``round'' of
CoD SnD for the purposes of this paper.

We note that SnD is not the only game mode played in competitive
CoD.\footnote{Hardpoint has been played as the first and fourth maps in
  a matchup. The variant for the third map was Domination in the 2020
  season, and Control in the 2021 and 2022 seasons.} Teams play in a
head-to-head, best-of-five format, where SnD is always played as the
second and fifth game modes. The best-of-five matchup could also be
called a ``series'', but since we analyze only the SnD games, we refer
to the SnD games as series.

@@

The data set consists of all SnD matches played in major tournaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852
series.\footnote{288 of the series occur in major tournaments, which are
  considered to be more competitive than the qualifiers since they are
  played in person (COVID-permitting), eliminating randomness due to
  online latency and ping. The qualifiers are played online.} Data has
been collected in spreadsheets by community member
``IOUTurtle''.\footnote{Raw data: \url{https://linktr.ee/CDLArchive}.
  Author: \url{https://twitter.com/IOUTurtle}. Our clean data can be
  found at
  \url{https://github.com/tonyelhabr/fps_round_games/blob/master/data/cod_rounds.csv}.}

6-0 sweeps make up 4.7\% of all series, while 6-5 series make up 23.7\%
of series. The observed offensive round win percentage across all rounds
is \(\bar{\tau}_O\) = 47.8\%.\footnote{Offensive round win percentage
  has been nearly constant across the three games during the CDL era: 1.
  47.2\% in MW (2020) 2. 47.9\% in Cold War (2021) 3. 48.1\% in Vanguard
  (2022)} Table \ref{tbl:o-win-prop-by-series-state} shows offensive
round win percentages by series ``state'', i.e.~the number of round wins
by each team prior to an upcoming round. Offensive round win percentage
is not quite constant, although never veers more than 10\% from
\(\bar{\tau}_O\).

@@

At tournaments, higher-seeded (``better'') teams choose whether they
want to start on offense in the SnD series, played as the second or
fifth game mode of the best-of-five matchup. Their opponent is assigned
to start on offense in the SnD slot not chosen.\footnote{Anecdotally,
  the higher-seeded teams tend to choose to play defense in the first
  round of the SnD series played as the second game mode in the
  best-of-five matchup, although this choice is not consistent across
  teams, or even with the same team over time.} Outside of tournaments,
teams all play each other twice throughout the season, constituting
qualifiers. Each team gets to play the higher-seeded role once in their
head-to-head matchups with a given team, despite their win-loss records.
Overall, we can argue that potential bias due to stronger teams playing
defense---or, offense, if offense happens to be generally advantageous
for a given map---is very minimal.

@@

There have been a handful of studies of the distribution of games played
in a series of a professional sport. Most assume a constant probability
\(\phi\) of a given team winning a game in the series, regardless of the
series state. Mosteller (1952) observed that the American League had
dominated the National League in MLB World Series matchups through 1952,
implying that games should not be modeled with a constant
\(\phi_0 = 0.5\). Mosteller proposed three approaches for identifying
the optimal constant probability value of the stronger team in the World
Series, finding \(\hat{\phi} \approx 0.65\) in each case: (1) solve for
\(\phi\) from the observed average number of games won by the loser of
the series, which he called the ``method of moments'' approach; (2)
maximize the likelihood that the sample would have been drawn from a
population in which the probability of a team winning a game is constant
across the series (i.e.~maximum likelihood), and (3) minimize the
chi-squared goodness of fit statistic \(\chi^2\) as a function of
\(\phi\).

Chance (2020) re-examined the constant probability notion in the MLB
World Series (1923--2018), the NBA Finals (1951--2018), and the NHL
Stanley Cup (1939--2018). Chance found strong evidence against the null
hypothesis of \(\phi_0 = 0.5\) in the MLB and NHL championship series
when applying Mosteller's first and second methods.

The concept of momentum goes hand-in-hand with a discussion of the
nature of series outcomes.\footnote{We often use ``streaks'' and
  momentum interchangeably, but as Steeger et al.~((2021)) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and the ``hot hand
fallacy'' (positive recency). Per Ayton et al. (2004), negative recency
is ``the belief that, for random events, runs of a particular outcome
\ldots{} will be balanced by a tendency for the opposite outcome'',
while positive recency is the expectation of observing future results
that match recent results.

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings, Gilovich et al. (1985),
henceforth GVT, did not find evidence for the hot hand phenomenon.
However, Miller and Sanjurjo (2018), henceforth MS, provided a framework
for quantifying streak selection bias, which effectively works in the
manner posited by the gambler's fallacy. Specifically, MS stated that a
``bias exists in a common measure of the conditional dependence of
present outcomes on streaks of past outcomes in sequential data''
implying that, under i.i.d. conditions, ``the proportion of successes
among the outcomes that immediately follow a streak of consecutive
successes is expected to be strictly less than the underlying
(conditional) probability of success''. When applying streak selection
bias to GVT's data, MS came to the opposite conclusions as GVT.

Other research has borrowed techniques from the field of quality
control, such as identifying unlikely sequences of events with the
Wald-Wolfowitz runs test (Peel and Clauset 2015, Steeger et al.~2021).
Peel and Clauset found no evidence for unlikely sequences of scoring
events in the NHL, College Football (CFB), and National Football League
(NFL), but did in the NBA. As a check on their entropy approach to
momentum identification, Steeger et al.~found several NHL teams with
sequence of wins in the 2018-2019 regular season that violated the
Wald-Wolfowitz null hypothesis.

@@

Despite the plethora of existing research on games played in a series
and momentum in sports, these topics have yet to be investigated
heavily, if at all, in esports. Work has been done to examine
intra-round, second-by-second win probability in other FPS titles such
as Counter-Strike (Xenopoulos, Freeman, and Silva 2022) and Valorant
(DeRover 2021), both of which are round-based like CoD SnD. However,
considering Counter-Strike and Valorant specifically, research on
round-level trends seems non-existent, perhaps for one of several
reasons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Both have economic aspects that can create clear advantages for one
  team in a round, given how prior rounds played out. CoD has no such
  equivalent, except for perhaps ``score streaks'', which infrequently
  occur.
\item
  Teams play either offense or defense for many consecutive rounds. (The
  former has a 15-15-1 format and the latter uses a 12-12-1 format.) On
  the other hand, teams in CoD SnD rotate roles every round, analogous
  to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series
  for professional sports like the MLB, NBA, and NHL.\footnote{1-1-1-1-1-1-1
    is not used today in these leagues, but it was at least once in
    each.} While theoretically one might be able to account for
  inter-round win percentage relationships in any kind of format, such
  as a 5-5-1, the rotation of team sides every round is convenient for
  convincing ourselves that rounds could reasonably be modeled as i.i.d.
  Bernoulli trials.
\item
  Both Counter-Strike and Valorant have overtime rules---team must win
  by two rounds---which can make end-of-series sequences difficult to
  model properly. CoD SnD does not have overtime rules.
\end{enumerate}

To our knowledge, there is no existing public statistical research on
the CoD SnD format, beyond descriptive analysis on social media. While
intra-round trends may be more directly applicable to teams looking for
an advantage on their competition, broader investigation of a concept
like the hot-hand fallacy in a sport where it has not yet been
invetigated---particularly one that is less subjected to factors that
may be difficult to control for, e.g.~weather---should be useful as a
precedent or supplement for future researchers.

@@

To evaluate the constant round win probability hypothesis, we can
compute the test statistic

\begin{equation}\label{eq:chi-squ}
\chi^2 = \sum_{r \in R} \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}
\end{equation}

in which \(\hat{\Phi}(r)\) and \(\Phi(r)\) represent the expected and
observed round win proportions respectively, and
\(R = [6, 7, 8, 9, 10, 11]\) for CoD SnD. The null hypothesis is that
the test statistic is chi-squared distributed, where the number of
degrees of freedom is \(\vert R \vert -1 = 5\) for Cod SnD. We reject
the null hypothesis if \(\chi^2 > \chi^2_{1-\alpha,\vert R \vert -1}\),
where \(\chi^2_{1-\alpha,\vert R \vert -1}\) is the chi-square critical
value with \(\vert R \vert -1\) degree of freedom and siginficance level
\(\alpha\).

@@

Let us now consider round win proportions immediately following a streak
of \(k\) wins, given that the series lasts \(r\) rounds. Removing our
knowledge of a streak, we might model the Bernoulli round win
probability as

@@

As implied by MS's Theorem 1, we should expect the proportion of round
wins immediately following a streak of \(k\) rounds wins for a series
lasting \(r\) rounds, \(\hat{P}^{+|k,r}_{MS}\), to be strictly less than
\(p^{+|r}\).\footnote{Unfortunately, there does not exist a closed form
  representation for the expected value of \(\hat{P}^{+|k,r}_{MS}\) for
  \(k > 1\). Nonetheless, one may run simulations to estimate the
  expected value, as we choose to do. We have adapted the code from Vafa
  (2017), which implements MS's framework (2018).}

Although one might be tempted to mirror the equal proportions hypothesis
testing performed by MS, our context is fundamentally different from
that of MS, who focus on longitudinal data in controlled settings,
unaffected by opposition.\footnote{GVT also perform statistical tests on
  shots from players in live games, i.e.~``observational'' data, but
  they note that their findings are likely affected player shot
  selection in the face of defensive strategy by the opposing team.} The
number of trials is fixed in their experimental designs, but in CoD SnD,
the number of rounds played is determined as a function of the max
number of possible rounds (\(s\)) and whether or not the team wins the
series. The Bernoulli trial success probability, i.e.~single round win
percentage in CoD SnD, is not independent of the opponent. Consequently,
a statistical test of the difference in \(\hat{P}^{+|k,r}\) and
\(\hat{P}^{-|kr}\), as performed by MS to evaluate their hypothesis
regarding post-streak success rate, is not completely appropriate,
although useful as a reference.

@@

We can perform a two-tailed binomial test to evaluate the null
hypothesis \(\omega = \omega_0\) for the observed probability of success
\(\omega\) and a user-specified \(\omega_0\) (where
\(0 \leq \omega_0 \leq 1\)). If there are \(r^+\) observed successes in
a sample of \(r\) trials and we expect that there should be
\(r * \omega_0\), the probability of arriving at this expected number of
successes is

@@

Treating the notional proportion \(\hat{P}^{+|k,r}_0\) as the null
\(\omega_0\) and plugging in the observed proportion \(P^{+|k,r}\) for
\(\omega\) in Equation \ref{eq:binom} (treating proportions as
probabilities), we can evaluate the null hypothesis that the observed
proportion of post-streak round wins is equal to the notional
proportion, given streak length \(k\) and series length \(r\). If we can
reject this null hypothesis, then we can consider team momentum,
represented by post-streak success, plausible.

As a reference, we can perform the same binomial test for the MS's
streak-selection-adjusted proportion, \(\hat{P}^{+|k,r}_{MS}\). However,
given the caveats mentioned before regarding applying MS's theorem to
our setting, the results of such binomial tests should be heeded with
caution.

@@

Again, we can apply a binomial test to evaluate the hypothesis that the
expected proportions, \(\hat{P}^{+|k,r,i}_0\) and
\(\hat{P}^{+|k,r,i}_{MS}\) separately, are equal to the observed
proportion \(P^{+|k,r}\).

@@

Stepping back from the (one-sided) perspective of a single team's round
win probability when streaking, one can attempt to detect the hot hand
phenomenon more broadly with a Wald-Wolfowitz runs test. Under the null
hypothesis, the number of runs in a sequence of \(r\) trials is a random
variable that can take on values \(+\) or \(-\) and arrive at \(r^+\)
successes (\(r^- = r - r^+\) failures), with the following mean \(\mu\)
and variance \(\sigma^2\):

@@

One can subset the observed series sequences, \(\zeta(r)\), to those
that violate the null hypothesis for the runs test and perform a test of
equal proportions, where the null is that the observed proportion of a
sequence relative to all possible sequences, \(P(\zeta) = P^\zeta\), is
equal to the expected proportion of the sequence relative to all
possible sequences, \(\hat{P}^\zeta\). The test statistic is

\begin{equation}\label{eq:prop}
z = \frac{P^\zeta - \hat{P}^\zeta}{\sqrt{P^\zeta_{\delta} (1 - P^\zeta_{\delta}) (1 / N^\zeta + 1 / \hat{N}^\zeta)  } }
\end{equation}

@@

and where \(N^\zeta\) and \(\hat{N}^\zeta\) are the observed and
expected number of sequences respectively. The null hypothesis is that
\(z\) is normally distributed. If \(z > +z_{(1-\alpha)/2}\) or
\(z < -z_{(1-\alpha)/2}\) for critical values at a significance level
\(\alpha\), we can reject the null hypothesis for such sequences and
build an argument in support of the hot hand effect.

@@

@@

Table \ref{tbl:mosteller-methods-results} shows the alternate values for
the constant round win probability that we find when applying the three
methods suggested by Mosteller (1952). Each is approximately equal to
0.575. When applying Equation \ref{eq:chi-squ}, each results in a
\(\chi^2\) value for which we cannot reject the constant probability
null hypothesis. Note that rejection of the null hypothesis is the
baseline and failure to reject the null is the non-trivial result,
contrary to most statistical analysis.

@@

Observing that the null hypothesis acceptance shown in Table
\ref{tbl:mosteller-methods-results}, along with the observation that
\(\hat{\Phi}_2(r)\) reasonably matches \(\Phi(r)\) (in comparison to
\(\hat{\Phi}_0(r)\)), we can say that the constant round win probability
assumption is valid in CoD SnD series with the appropriate choice of
\(\phi\) (\(\approx 0.575\)).

@@

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), we focus
on streaks of three round wins.\footnote{Three happens to also be a
  reasonable number for series that last at maximum 11 rounds.} Table
\ref{tbl:pwkr} compares the notional and MS expected proportions,
\(\hat{P}^{+|k,r}_0\) and \(\hat{P}^{+|k,r}_{MS}\) respectively, with
the observed round win proportion, \(P^{+|kr}\), following streaks of
\(k=3\) round wins given the series outcome.

@@

With the exception of \(\hat{P}^{+|k,r}_{MS}\) when \(r = 10\) and the
eventual series loser is the team that wins after a streak of three
round wins, all un-adjusted binomial test \(p\)-values are greater than
the \(\alpha = 0.05\) confidence level, implying that we cannot reject
the null hypothesis that the expected notional and MS post-streak round
win proportions are different than the observed proportion. However,
after adjusting the \(p\)-values with the Benjamini and Yekutieli (BY)
correction (2001), the null hypothesis cannot be rejected for any
case.\footnote{As noted before, the MS expectations may be unreliable
  for CoD SnD, so one is inclined to prefer the results of the notional
  binomial tests.}

@@

However, our results do not provide evidence for momentum on several
fronts: (1) the evidence supporting the constant round probability; (2)
the failure to reject the binomial null hypothesis for post-streak win
proportions; and (3) the failure to reject the equal proportions null
hypothesis for streaky round sequences identified by the Wald-Wolfowitz
runs test.
